Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-17 ICLR25/formula.tex
updated
update the section of formula.tex

	Let $a$ be a node in the network (at least the second hidden layer) and we have computed a series of lower and upper bound for all nodes before the layer of $a$. If we consider to compute the upper bound of $a$ by the standard MILP model.

	\begin{enumerate}
		\item $M^a_X$ means the MILP model with target node $a$ and open node set $X$. $M^a_{\emptyset}$ is the LP model for $a$.

		\item For a set $X$, if the upper bound of $a$ computed from $M^a_{\emptyset}$ is $u_0$, and from $M^a_X$ is $u_X$, then the improvement (of upper bound of $a$) of $X$, $I_X$ (or $I_X^{a,u}$) is defined as : $$I_X = u_0-u_X.$$ 
	\end{enumerate}
\end{definition}

Similarly we can define the improvement of lower bound to be $I_X^{a,l} = l_X-l_0$.

@@

\begin{align*}
	|(I_X - \sum_{b\in X} I_b)/I_X| < 1\%. \ (\text{in experiments})
\end{align*} And even $X$ comes from 3 layers before the target layer, in experiments, $I_X$ is still close to $\sum_{b\in X} I_b$.

Therefore, based on this observation, the question to choose $X$ is convert to compute $I_b$ for nodes $b$ in 3 layers before the target layer.

@@

The formula to compute the improvement of a single node is the actual content of our formula. For different layers, the formula will be different.

For one layer before the target layer, it is very simple and more accurate. Notice that, either the observation in last subsection, or formula in this subsection, does not have solid theoretical proof to show they are very accurate. They are all based on experiments.

Suppose $b$ is a node one layer before $a$. To compute $I_b$, we need to run $M^a_{\emptyset}$ to compute the upper bound of $a$ first to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

\begin{definition}
	Let $v(b)$ be the value of $b$ in the LP solution of upper bound of $a$, and $v(\hat{b})$ be the value of $\hat{b}$.

	Then the formula to estimate improvement of upper bound of $b$ is: \begin{align*}
		I_b \approx W_{ba}(v(\hat{b})-\ReLU(v(b))).
	\end{align*}

	Specially, if $v(\hat{b})=\ReLU(v(b))$, then \begin{align*}
		I_b = W_{ba}(v(\hat{b})-\ReLU(v(b))) = 0.
	\end{align*}

	Similarly, let $v(b)$ be the value of $b$ in the LP solution of lower bound of $a$, and $v(\hat{b})$ be the value of $\hat{b}$.

	Then the formula to estimate improvement of lower bound of $b$ is: \begin{align*}
		I_b \approx -W_{ba}(v(\hat{b})-\ReLU(v(b))).
	\end{align*}

@@

To explain the formula, we use upper bound and $W_{ba} > 0$ as an example. To compute the upper bound of $a$, $\hat{b}$ should be as large as possible. In the LP model, for fixed $v(b)$, the upper bound of $\hat{b}$ may be larger than $\ReLU(v(b))$. This is because in LP model, the upper bound of $v(\hat{b})$ is decided by the linear approximation rather than $\ReLU$ function. So, then node $b$ is open, if values before $v(\hat{b})$ do not change, the the upper bound of $a$ will improve by lower $v(\hat{b})$ to $\ReLU(v(b))$.

Of course changing other variables may effect the upper bound, but our experiments show that, the change from $v(\hat{b})$ to $\ReLU(v(b))$ is the major source of improvement.

For earlier layers, the formula will be more complex and less accurate. But we will still focus on the source $v(\hat{b})-\ReLU(v(b))$.

@@

Basically, we hope to use $\sum_c W_{ca}W_{bc}$ to replace the coefficient $W_{bc}$ in last subsection.

However, directly replace $W_{ba}$ with the new coefficient will lose too much accuracy. So we try to improve the accuracy by considering more details from the LP solutions.

For nodes $c$ in the layer between $a$ and $b$, according to the LP solution, we may check $v(c)$ is positive or negative. And as we discussed in last subsection, $W_{ca}$ can decide whether $v(\hat{c})$ is more close $\ReLU(v(c))$ or the linear approximation. That is the following definition.

\begin{definition}
	For a fixed target node $a$ and a fixed source node $b$ in two layers before $a$. To compute the improvement of of upper bound of $a$ by $b$, we define the following function for nodes in 1 layer before $a$:
	\begin{align*}
		k(c) =
		\begin{cases}
			1, & \text{if } v(c) > 0 \text{ and } W_{ca} < 0\\
			0, & \text{if } v(c) \leq 0 \text{ and } W_{ca} < 0\\
			\frac{\UB(c)}{\UB(c)-\LB(c)}, & \text{if }W_{ca} > 0
		\end{cases}
	\end{align*} and use the following formula to compute the estimation of improvement: \begin{align*}
	I_b = (v(\hat{b})-\ReLU(v(b)))\sum_c W_{bc}W_{ca}k(c)
	\end{align*}

However, the error is still too large. This is because the improvement of a node $c$ can not exceed its precomputed upper and lower bound (we may use $UB(c)$ and $LB(c)$ to denote them). Considering this, we may improve our formula to the following:

\begin{definition}
	For a fixed target node $a$ and a fixed source node $b$ in two layers before $a$. Let $\UB$ and $\LB$ denote the precomputed upper bounds and lower bound used in building MILP models. To compute the improvement of of upper bound of $a$ by $b$, we use the following formula:

	We define a distance function $D$:

	\begin{align*}
		&D(b) = \ReLU(v(b))-v(\hat{b})\\
			&D(c) =
		\begin{cases}
			\max(\LB(c), 0, v(c)+D(b)k(c)W_{bc})-v(\hat{c}), & \text{if }W_{ca} < 0\\
			\min(\UB(c), v(\hat{c})+D(b)k(c)W_{bc})-v(\hat{c}), & \text{if }  W_{ca} > 0
		\end{cases}\\
		&D(a) = -\sum_c D(c)W_{ca}
	\end{align*}
\end{definition}

This is the formula to compute improvement of nodes two layers before.

@@

This formula is based on previous subsection but more complex. In some network, run this formula may cost too much time.

The key problem is that, how to compute the coefficient $k$ for nodes in 2 layers before the target layer. We use the following method to estimate the coefficient.

@@

	\begin{align*}
		&w_0 = \sum_C D(c)W_{cd}\\
		&w_1 = \min(\UB(d),v(d)+w_0)\\		
		&D(d) =
		\begin{cases}
			w_1-v({d}), & \text{if }W_{da} > 0 \text{ and } \LB(d)\geq 0\\
		k(d)(w_1-v({d})), & \text{if }W_{da} > 0 \text{ and } \LB(d)< 0\\
		\ReLU(w_1)-v(\hat{d})	, & \text{if }  W_{da} < 0
		\end{cases}\\
		&D(a) = -\sum_c D(d)W_{da}
	\end{align*}
\end{definition}

>>>>

	Let $a$ be a node in the network (in the second hidden layer or deeper), and we have computed a series of lower and upper bound for all nodes before the layer of $a$. If we consider to compute the upper bound or lower bound of $a$ by the standard MILP model, then

	\begin{enumerate}
		\item Let $M^a_X$ mean the MILP model with target node $a$ and open node set $X$. Hence $M^a_{\emptyset}$ is the LP model for $a$.

		\item For a set $X$, if the upper bound of $a$ computed from $M^a_{\emptyset}$ is $u_0$, and from $M^a_X$ is $u_X$, then the improvement (of upper bound of $a$) of $X$, $I_X^{a,u}$ (or $I_X$ if there is no risk of confusion) is defined as : $$I_X^{a,u} = u_0-u_X.$$ Similarly we can define the improvement of lower bound to be $I_X^{a,l} = l_X-l_0$.
	\end{enumerate}
\end{definition}

@@

\begin{align}
	|(I_X - \sum_{b\in X} I_b)/I_X| << 1. \ (\text{in experiments})
\end{align} Even $X$ contains nodes from 3 layers before the target layer, in experiments, $I_X$ is still close to $\sum_{b\in X} I_b$.

Therefore, based on this observation, the question to choose $X$ is converted to compute $I_b$ for nodes $b$ in layers before the target layer. Our formula is to estimate the improvement of different individual nodes in different layers. For different layers, the formula will be different.  However, neither the observation in this subsection nor the formula in the next subsection has solid theoretical proof to show that they are very accurate. They are all based on experiments.

@@

For one layer before the target layer, it is very simple and most accurate.

Suppose $b$ is a node in one layer before $a$. To estimate $I_b$, first we need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

\begin{definition}
	Let $v(b)$ be the value of $b$ in the LP solution of upper bound of $a$, and $v(\hat{b})$ be the value of $\hat{b}$.	Then the formula to estimate improvement of upper bound of $b$ is: \begin{align*}
		I_b \approx W_{ba}(v(\hat{b})-\ReLU(v(b))).
	\end{align*}

	Specially, if $v(\hat{b})=\ReLU(v(b))$, then we can prove \begin{align*}
		I_b = W_{ba}(v(\hat{b})-\ReLU(v(b))) = 0.
	\end{align*}

	Similarly, let $v(b)$ be the value of $b$ in the LP solution of lower bound of $a$, and $v(\hat{b})$ be the value of $\hat{b}$. Then the formula to estimate improvement of lower bound of $b$ is: \begin{align*}
		I_b \approx -W_{ba}(v(\hat{b})-\ReLU(v(b))).
	\end{align*}

@@

To explain the formula, we use upper bound and the case that $W_{ba} > 0$ as an example. To compute the upper bound of $a$, $\hat{b}$ should be as large as possible. In the LP model, for fixed $v(b)$, the upper bound of $\hat{b}$ may be larger than $\ReLU(v(b))$. This is because in LP model, the upper bound of $v(\hat{b})$ is decided by the linear approximation rather than $\ReLU$ function. So, when node $b$ is open, if $v(b)$ do not change, then the upper bound of $a$ will be improved because the value of $v(\hat{b})$ will be lower to $\ReLU(v(b))$.

Of course changing other variables may also effect the upper bound, but our experiments show that, the change from $v(\hat{b})$ to $\ReLU(v(b))$ is the major source of improvement.

@@

For earlier layers, the formula will be more complex and less accurate. We will still focus on the source $v(\hat{b})-\ReLU(v(b))$. Basically, we hope to use $\sum_c W_{ca}W_{bc}$ to replace the coefficient $W_{bc}$ in last subsection. However, directly using $\sum_c W_{ca}W_{bc}$ will lose too much accuracy. So we try to improve the accuracy by considering more details from the LP solutions.

For nodes $c$ in the layer between $a$ and $b$, according to the LP solution, we may check $v(c)$ is positive or negative. Next we consider the following definition.

\begin{definition} \label{2layer}
	Suppose we have a fixed target node $a$ and a fixed source node $b$ in two layers before $a$. Let $\UB,\LB$ be the function of precomputed upper and lower bound before the layer of $a$, and let $v$ be the function of values of nodes in the LP solution of upper bound of $a$. 
	To compute the improvement of of upper bound of $a$ by $b$, we define the following function for nodes $c$ in one layer before $a$:
	\begin{align*}
		k(c) =
		\begin{cases}
			1, & \text{if } v(c) > 0 \text{ and } W_{ca} < 0\\
			0, & \text{if } v(c) \leq 0 \text{ and } W_{ca} < 0\\
			\frac{\UB(c)}{\UB(c)-\LB(c)}, & \text{if }W_{ca} > 0
		\end{cases}
	\end{align*} 
\end{definition}

We may use the following formula to estimate the improvement:  \begin{align*}
	I_b \approx (v(\hat{b})-\ReLU(v(b)))\sum_c W_{bc}W_{ca}k(c).
\end{align*}However, the error is still too large. This is because the improvement of a node $c$ can not exceed its precomputed upper and lower bound. Considering this, we may improve our formula.

\begin{definition}
	Continue the assumption in Definition \ref{2layer} To compute the improvement of of upper bound of $a$ by $b$, we define a distance function $D$:
	\begin{align*}
		&D(b) = \ReLU(v(b))-v(\hat{b})\\
			&D(c) =
		\begin{cases}
			\max(\LB(c), 0, v(c)+D(b)k(c)W_{bc})-v(\hat{c}), & \text{if }W_{ca} < 0\\
			\min(\UB(c), v(\hat{c})+D(b)k(c)W_{bc})-v(\hat{c}), & \text{if }  W_{ca} > 0
		\end{cases}\\
		&D(a) = -\sum_c D(c)W_{ca}
	\end{align*}
\end{definition}

Then $D(a)$ is the estimation of improvement of upper bound of $a$ by opening node $b$. This is the formula to compute improvement of nodes two layers before.

@@

This formula is based on previous subsection but more complex. In some network, run this formula may cost too much time but obtain little accuracy.

The key problem is how to compute the coefficient $k$ for nodes in two layers before the target layer. We use the following method to estimate the coefficient.

@@

	\begin{align*}
		&w_0 = \sum_C D(c)W_{cd}\\
		&w_1 = \min(\UB(d),v(d)+w_0)\\		
		&D(d) =
		\begin{cases}
			w_1-v({d}), & \text{if }W_{da} > 0 \text{ and } \LB(d)\geq 0\\
		k(d)(w_1-v({d})), & \text{if }W_{da} > 0 \text{ and } \LB(d)< 0\\
		\ReLU(w_1)-v(\hat{d})	, & \text{if }  W_{da} < 0
		\end{cases}\\
		&D(a) = -\sum_c D(d)W_{da}
	\end{align*}
\end{definition} $D(a)$ is the estimation of improvement of upper bound of $a$ by opening node $b$.
